Noun Serialization

### 2025-12-27 mss.tex
Post
Fix
Revise

\usepackage{ustj}
\usepackage{cancel}
\usepackage{pdflscape}

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).}
\label{tab:atom-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b110} \\
  \texttt{0b10} & 2 & \texttt{0b10100} \\
  \texttt{0b11} & 2 & \texttt{0b11100} \\
  \texttt{0b101} & 3 & \texttt{0b1011000} \\
  \texttt{0b111} & 3 & \texttt{0b1111000} \\
  \texttt{0x70} & 7 & \texttt{0b111000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
  \texttt{0x1234} & 13 & \texttt{0b1001000110100} \\
  & & \texttt{↳10000000000000} \\
\end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings (\texttt{+fat} encoding).  Elided digits are shown in light gray for clarity.}
\label{tab:fat-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b\textcolor{lightgray}{1}10} \\
  \texttt{0b10} & 2 & \texttt{0b\textcolor{lightgray}{10}100} \\
  \texttt{0b11} & 2 & \texttt{0b\textcolor{lightgray}{1}1100} \\
  \texttt{0b101} & 3 & \texttt{0b\textcolor{lightgray}{10}11000} \\
  \texttt{0b111} & 3 & \texttt{0b\textcolor{lightgray}{1}111000} \\
  \texttt{0x70} & 7 & \texttt{0b\textcolor{lightgray}{1}11000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
  \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}1000110100} \\
  & & \texttt{↳10000000000000} \\
\end{tabular}
\end{table}

@@

Consider the noun \texttt{[a [b c]]} (= \texttt{[a b c]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:

@@

\begin{table}[ht]
\centering
\caption{Example noun encodings (\texttt{+gel} encoding).  \texttt{.} dot delimits bit fields for readability.}
\label{tab:gel-encodings}
\begin{tabular}{ r r }
  \multicolumn{1}{l}{\textbf{Noun}} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
  \hline
  \texttt{0} & \texttt{0b0.1.0} \\
  \texttt{[0 0]} & \texttt{0b1.0.1.0.1} \\
  \texttt{[1 0]} & \texttt{0b1.0.110.0.1} \\
  \texttt{[2 1 0]} & \texttt{0b1.0.110.0.1.100100.0.1} \\
  \texttt{[[2 3] 1 0]} & \texttt{0b1.0.110.0.1.110100.0.100100.0.1.1} \\
\end{tabular}
\end{table}

@@

\noindent
Decoding is left as an exercise to the interested reader.

@@

\selectlanguage{USenglish}
% \printbibliography

\section*{Appendix:  Python \texttt{+jam}/\texttt{+cue}}
\subsection*{Paul Driver \patp{fodwyt-ragful}}

The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}

@@

    def save(a: noun):
        refs[a] = cur
    def mat(i: int):
        if 0 == i:
            one()
        else:
            a = i.bit_length()
            b = a.bit_length()
            above = b + 1
            below = b - 1
            bits(1 << b, above)
            bits(a & ((1 << below) - 1), below)
            bits(i, a)

>>>>

\usepackage{ustj}
\usepackage{cancel}
\usepackage{pdflscape}
\usepackage{array}  % for table column formatting

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).  Elided digits are shown in light gray for clarity.}
\label{tab:atom-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b\textcolor{lightgray}{0}1} \\
  \texttt{0b1} & 1 & \texttt{0b110} \\
  \texttt{0b10} & 2 & \texttt{0b10100} \\
  \texttt{0b11} & 2 & \texttt{0b11100} \\
  \texttt{0b101} & 3 & \texttt{0b1011000} \\
  \texttt{0b111} & 3 & \texttt{0b1111000} \\
  \texttt{0x70} & 7 & \texttt{0b111000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
  \texttt{0x1234} & 13 & \llap{\texttt{0b100100011010010000000000000}} \\
\end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings (\texttt{+fat} encoding).  Elided digits are shown in light gray for clarity.}
\label{tab:fat-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b\textcolor{lightgray}{1}10} \\
  \texttt{0b10} & 2 & \texttt{0b\textcolor{lightgray}{10}100} \\
  \texttt{0b11} & 2 & \texttt{0b\textcolor{lightgray}{1}1100} \\
  \texttt{0b101} & 3 & \texttt{0b\textcolor{lightgray}{10}11000} \\
  \texttt{0b111} & 3 & \texttt{0b\textcolor{lightgray}{1}111000} \\
  \texttt{0x70} & 7 & \texttt{0b\textcolor{lightgray}{1}11000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
  \texttt{0x1234} & 13 & \llap{\texttt{0b\textcolor{lightgray}{100}100011010010000000000000}} \\
  % \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}100011010010000000000000} \\
  % \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}1000110100} \\
  % & & \texttt{↳10000000000000} \\
\end{tabular}
\end{table}

@@

Consider the noun \texttt{[4 [0 1]]} (= \texttt{[4 0 1]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:

@@

\begin{table}[ht]
\centering
\caption{Example noun encodings (\texttt{+gel} encoding).  \texttt{.} dot delimits bit fields for readability.}
\label{tab:gel-encodings}
\begin{tabular}{ r r }
  \multicolumn{1}{l}{\textbf{Noun}} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
  \hline
  \texttt{0} & \texttt{0b\textcolor{lightgray}{0}.1.0} \\
  \texttt{[0 0]} & \texttt{0b1.0.1.0.1} \\
  \texttt{[1 0]} & \texttt{0b1.0.110.0.1} \\
  \texttt{[2 1 0]} & \texttt{0b1.0.110.0.1.100100.0.1} \\
  \texttt{[[2 3] 1 0]} & \texttt{0b1.0.110.0.1.110100.0.100100.0.1.1} \\
\end{tabular}
\end{table}

@@

\noindent
Decoding is left as an exercise to the motivated reader.

@@

\selectlanguage{USenglish}
% \printbibliography
\newpage

\section*{Appendix:  Python \texttt{+jam}/\texttt{+cue}}
\subsection*{Paul Driver \patp{fodwyt-ragful}}
% add to table of contents
\addcontentsline{toc}{section}{Appendix:  Python \texttt{+jam}/\texttt{+cue}}

The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.%\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}

@@

    def save(a: noun):
        refs[a] = cur

    def mat(i: int):
        if 0 == i:
            one()
        else:
            a = i.bit_length()
            b = a.bit_length()
            above = b + 1
            below = b - 1
            bits(1 << b, above)
            bits(a & ((1 << below) - 1), below)
            bits(i, a)
